---
title: "Replicating Nguyen (2009) with `CIPerm`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nguyen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We show that our package replicates the analyses in Nguyen (2009).


First, input the data from Table 1:

```{r setup}
library(CIPerm)
x <- c(19, 22, 25, 26)
y <- c(23, 33, 40)
```

Replicate Table 2 along with the three extra columns from Table 3:

```{r table2}
demo <- dset(x, y)
knitr::kable(demo)
```

Replicate the left-tailed p-values reported on pages 5-6.  
The first three should be `0.08571429`, and the last should be `0.1142857`.

```{r pval}
# Difference in means
pval(demo, "Left", "m")

# Sum of treatment group
pval(demo, "Left", "s")

# Difference in medians
pval(demo, "Left", "d")

# Wilcoxon rank sum statistic
pval(demo, "Left", "w")
```

Replicate the confidence interval (left-tailed, with confidence level $(1 - \frac{2}{35})\times 100\% \approx 94.3\%$) reported on page 11.  
This should be `c(-Inf, 2)`.

```{r cint}
cint(demo, 2/35, "Left")
```

